Forecasting hourly hospital admission

### 2021-10-26 paper/paper.Rmd
JB paper additions

We develop a range of models for forecasting hospital admissions using a multi-year dataset. In order to facilitate model development and testing, we divide the data into a training period. We use cross-validation to iteratively fit, evaluate and develop models using the training data, before settling on their optimal configuration and testing them on the as-yet unseen test data.

Forecast are produced/issue at midnight and mid-day for lead-times from 0 to 48 hours-ahead on a rolling basis. In order to compare both parametric and non-parametric probabilistic forecasts, we evaluate quantiles of the predictive distribution at probability level 0.05, 0.10,... 0.95, as well as the expectation.

## data {#data}

The data we use in this case study are from hospital XYZ, and comprises records of arrivals at Accident and Emergentcy at hourly granularity. The training data runs from 01/04/2014 to 28/02/2018, and the test data runs from 1/3/2018 to 28/2/2019.

@@

However, experiments on the training data reveal that calibration of forecasts based on the Poisson distribution is poor, suggesting that the shape of the distribution is unsuitable for the present application. In particular, we observe that forecast uncertainty appears to vary depending on the time of day and possibly other explanatory variables. Therefore, we consider more flexible, two parameter distributions so that we are able to specify additive models for both location and scale parameters, specifically the truncated Normal distribution, with truncation at 0. The resulting density forecasts are given by
\begin{equation}
  F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)}
 (\#eq:truncatedn)
\end{equation}
with additive models
\begin{align*}
  \mu_t = &  \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\
  log(\sigma_t) & = \sum_{i=1}^10 D^{+}_i(t) f(H(t))
\end{align*}
for the mean and variance parameters. We also performed experiments with the truncated $t$ distribution and negative binomial distribution, but these did not results in forecasts as well calibrated as the truncated normal.

@@

The objective of density forecasts it to be as sharp as possible while remaining reliable/calibrated [@Gneiting2007a]. A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, witch is valuable to decision makers provided the forecast is calibrated.

The Pinball Score is a strictly proper score used to evaluate quantile forecasts and is the discrete form of the Continuous Rank Probability Score. It rewards sharpness and penalises mis-calibration, so measures all-round performance, however, calibration should still be verified separately. Furthermore, 
The Pinball Score for an individual quantile matches the loss function minimised in quantile regression model estimation. The Pinball Score is given by

@@

Calibration, also called `reliability`, is the property that forecast probabilities match the observed frequency of realisations. If a forecast is calibrated, then $20\%$ of observations should fall below the $\alpha=0.2$ predictive quantile (with some tolerance based on the finite sample size). This property is necessary for forecast probabilities to be used in quantitative decision-making. Calibration is typically evaluated visually using reliability diagrams, which plot the nominal coverage, $\alpha$, against observer frequency mean($\mathbb{1}(y_{t}\le q_{\alpha,t})$).

@@

## Case study

The data is portioned into training and test data, with all model development and hyper-parameter tuning performed using training data only. Data from 2014-04-01 to 2018-02-28 are used for training, and from 2018-03-01 to 2019-02-28 for testing.

@@

**Not sure if this should stay here, but interesting and worth reporting, I think...**

Plot smooth functions estiamted for time of day, day of year, for both mean and variance parameters.

>>>>

We develop a range of models for forecasting hospital admissions using a multi-year dataset from a hospital in the UK. In order to facilitate model development and testing, we divide the data into a training and test period. We use cross-validation to iteratively fit, evaluate and develop models using the training data, before settling on their optimal configuration and testing them on the as-yet unseen test data. Data from 2014-04-01 to 2018-02-28 are used for training, and from 2018-03-01 to 2019-02-28 for testing.

Forecast are issued every day at midnight and mid-day for lead-times from 0 to 48 hours-ahead on a rolling basis. In order to compare both parametric and non-parametric probabilistic forecasts, we evaluate quantiles of the predictive distribution at probability level 0.05, 0.10,... 0.95, as well as the predictive expectation.

In the remainder of this section we explore the data in more detail, describe the forecast evaluation metrics we will employ, and introduce a range of benchmark models for producing forecasts. The benchmarks serve as reference methods and include some simple methods as well as more sophisticated routines designed to work `out-of-the-box', such as Facebook's Prophet. The performance of these methods provides an important reference, and any proposed novel solution should offer material benefits to justify increased complexity.

## Data exploration {#data}

The data we use in this case study are from hospital XYZ, and comprises records of arrivals at Accident and Emergency at hourly granularity. The training data runs from 01/04/2014 to 28/02/2018, and the test data runs from 1/3/2018 to 28/2/2019.

@@

However, experiments on the training data reveal that calibration of forecasts based on the Poisson distribution is poor, suggesting that the shape of the distribution is unsuitable for the present application. In particular, we observe that forecast uncertainty appears to vary depending on the time of day and possibly other explanatory variables. Therefore, we consider more flexible, two parameter distributions so that we are able to specify additive models for both location and scale parameters, specifically the truncated Normal distribution, with truncation at 0. The resulting density forecasts are given by
\begin{equation}
F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)}
(\#eq:truncatedn)
\end{equation}
with additive models
\begin{align*}
\mu_t = &  \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\
log(\sigma_t) & = \sum_{i=1}^10 D^{+}_i(t) f(H(t))
(\#eq:additivemodel_NOtr_v2)
\end{align*}
for the mean and variance parameters. We also performed experiments with the truncated $t$ distribution and negative binomial distribution, but these did not results in forecasts as well calibrated as the truncated normal.

@@

The objective of density forecasts it to be as sharp as possible while remaining reliable/calibrated [@Gneiting2007a]. A forecast is said to be sharp if the predictive distribution has a relatively small spread, indicating low uncertainty, which is valuable to decision makers provided the forecast is calibrated.

The Pinball Score is a strictly proper score used to evaluate quantile forecasts and is the discrete form of the Continuous Rank Probability Score. It rewards sharpness and penalises mis-calibration, so measures all-round performance, however, calibration should still be verified separately. Furthermore, the Pinball Score for an individual quantile matches the loss function minimised in quantile regression model estimation. The Pinball Score is given by

@@

Calibration, also called `reliability', is the property that forecast probabilities match the observed frequency of realisations. If a forecast is calibrated, then $20\%$ of observations should fall below the $\alpha=0.2$ predictive quantile (with some tolerance based on the finite sample size). This property is necessary for forecast probabilities to be used in quantitative decision-making. Calibration is typically evaluated visually using reliability diagrams, which plot the nominal coverage, $\alpha$, against observer frequency mean($\mathbb{1}(y_{t}\le q_{\alpha,t})$).

@@

@@

We can support the interpretation the models given in Equation \@ref(eq:additivemodel_NOtr_v2) after the parameters have been estimated by visualising their reposes to inputs. Figure \@ref(fig-NOtr-model) shows the effect of time of day and day-type on the level and spread of predicted arrivals.

```{r fig-NOtr-model}

load("NOtr2_model.Rda")

plot_data <- data.frame(Hour = 0:23)
for(i in c("Mon","Tue","Wed","Thu","Fri","Sat","Sun","Public Holiday","Christmas Day","New Years Day")){
  pef <- gamlss::getPEF(h2_gamlss2,term="clock_hour",parameter = "mu",fixed.at = list(dow3=i))
  plot_data[[i]] <- pef(0:23)
}

plot_data <- reshape::melt.data.frame(plot_data,id.vars="Hour",variable_name="Day Type")

plot_1 <- ggplot(plot_data,aes(x=Hour,y=value,color=`Day Type`)) + geom_smooth(se = F,span=0.5) +
  ylab("Effect [Expected Arrivals]") + theme_few()

# plot_1

plot_data2 <- data.frame(Hour = 0:23)
for(i in c("Mon","Tue","Wed","Thu","Fri","Sat","Sun","Public Holiday","Christmas Day","New Years Day")){
  pef <- gamlss::getPEF(h2_gamlss2,term="clock_hour",parameter = "sigma",fixed.at = list(dow3=i))
  plot_data2[[i]] <- pef(0:23)
}

plot_data2 <- reshape::melt.data.frame(plot_data2,id.vars="Hour",variable_name="Day Type")

plot_2 <- ggplot(plot_data2,aes(x=Hour,y=value,color=`Day Type`)) + geom_smooth(se = F,span=0.5) +
  ylab("Effect [Standard Deviation of Arrivals]") + theme_few()

#plot_2

grid.arrange(plot_1, plot_2, ncol=2)
